---
title: "Graphics and visualization"
author: "Week 3"
date: "07/15/2030-07/16/2020"
output: 
  html_document:
    toc: true
    toc_float: true
---


## Goal and Video

The goal of this lab is to become familiar with the grammar of graphics 
philosophy implemented in the `R` package `ggplot2`. This is a newer version than basic R plots. However, it is a more powerful tool, and completely beginner friendly, so we won't go through the basic R-plots and will get you accustomed right away to this particular way of making plots.
- __To do for Thursday__ Begin by watching the  second part of the [lecture on data visualization](https://drive.google.com/file/d/1BBNvt2EWtZnixHbGxnXg-vbivP7Iu0EI/view?usp=sharing) (57 min to the end).
 
 
- Work through the second part of this lab by running all the R code to your computer and making sure that you understand the input and the output. 



## Getting Started (Wednesday)

We will be going through some of the topics from the `ggplot2` documentation
website, [docs.ggplot2.org](http://docs.ggplot2.org). You should keep that website open as 
a reference throughout this tutorial.

Another useful resource, the complete guide to the Grammar of Graphics 
philosophy, is available through [Springer website](https://link.springer.com/book/10.1007%2F978-3-319-24277-4) 
(to which you have a free legal access with your Stanford ID).

### Required Packages (Wednesday)

In order to learn how to use `ggplot2`, we will use a few datasets 
available through the `nutshell`, `parathyroidSE`, `EnsDb.Hsapiens.v86` packages.

The code below checks if all packages required for today's lab are available,
and installs any missing ones. Note that (in the template file) 
the chunk declaration includes the following options: 
`message = FALSE, warning = FALSE, results = "hide"`. 
These R Markdown settings let you suppress messages and warnings which would 
otherwise appear below the chunk. Another way to avoid the lengthy messages
during package loading, is to wrap R commands 
inside `suppressMessages()` function, e.g. `suppressMessages(library("foo"))`.

```{r setup, message = FALSE, warning = FALSE, results = 'hide'}
pkgs_needed <- c("tidyverse", "Biostrings", 
                 "parathyroidSE", "EnsDb.Hsapiens.v86", "nutshell")
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  BiocManager::install(letsinstall)
}
library("dplyr")
library("ggplot2")
library("Biostrings")
```


### Load Data

First, we will use a data set about births in the USA in 2006. You can read more 
about this in the *"R in a Nutshell, 2nd edition"* book which is freely available 
as a PDF file online.  This is convenient since you can follow the same methods 
in that book (except translating into the grammar of graphics to make plots). You can download the dataset [here](https://biox-rbootcamp.github.io/biox-rbootcamp.github.io/assets/data/births2006.smpl.rda) from the course's website.

First, we load the dataset and add two new variables: 
one to encode whether the day falls on a weekend or whether it is a weekday,
and a three-level, discretized health score based on the APGAR 5 score.

We also fix an ideosyncratic encoding of not available values by 99 with proper
`NA`s. 

Finally, we define a subsampled version of the dataset `births_small` to
speed up some of the plotting (feel free to try out the computations and plots
we see subsequently with the full dataset `births`.)

```{r births}
load("~/Downloads/births2006.smpl.rda")

births <- mutate(births2006.smpl,
  WEEKEND = ifelse(DOB_WK %in% c(1, 7), "Weekend", "Weekday"),
  HEALTH  = c("CritLow", "Low", "Normal")[ 1 +
    findInterval(APGAR5, c(3.5, 6.5)) ],
  ESTGEST = replace(ESTGEST, ESTGEST==99, NA))

set.seed(1)
births_small <- births[ sample(nrow(births), 40000), ]
head(births_small)
```


## Plot initialization & `geom`'s

> ggplot2 is a plotting system for R, based on the grammar of graphics. 
It takes care of many of the fiddly details that make plotting a hassle 
(like drawing legends) as well as providing a powerful model of graphics 
that makes it easy to produce complex multi-layered graphics [^1]

[^1]: http://ggplot2.org/

You can think of plots generated with `ggplot2` as sentences.
The function **`ggplot()` starts a sentence (initializes a plot)** which usually 
contains a noun/subject (a dataset with information to be plotted). 

The `+` operator is used to add further clauses/fragments containing "verbs", 
"adjectives", "adverbs" etc. to the sentence. This allows you to construct 
sophisticated plots from a few elementary building blocks, just like
forming compound sentences from simple phrases. 

In `ggplot` world, verbs are **geometric objects, or `geom`'s**.
For example, `geom_bar()` draws bars, `geom_point()` draws points. 
There can be multiple `geom`s in a sentence, and these are then also called 
layers. 

For each `geom`, you need to define required **aesthetics** 
using the function `aes()`. The aesthetics of a `geom` determine
how the attributes of the geometric object (e.g. the x- and y- coordinates of 
the points, their colors and shapes, etc.) are mapped to the columns of the
supplied data.frame. The aesthetics mapping can be defined when initializing
a plot (with `ggplot(data = dataframe_name, aes(...))`), which makes it apply 
to all `geom`'s. Otherwise, one can specify `aes()` mapping for each `geom` 
separately (e.g. ` geom_bar(aes(...))`), in which case it applies to that 
particular `geom` only.

It is also possible to use a different dataset for `geom` than the one supplied
to `ggplot()`. To do this you simply supply a new data.frame as a `data` 
argument of `geom` (e.g. `geom_point(data = new_dataframe_name, aes(...))`)
which would overwrite the dataset used to this `geom`.


For example, let's look at the number of births on each day of the week in 
our sample data. First set up our plot with the data `births`.

```{r birth_init_ggplot}
ppp <- ggplot(births_small) 
```

Note that this doesn't actually plot anything (just like if you wrote 
`junk <- 7` doesn't output anything until you run `junk`.
What happens here when you run `ppp`? 

The number of births per day of the week can be shown easily in a barplot,
so let's use that. To create a geometric object layer for a **barplot we use the 
function `geom_bar()`**. In order for it to know what part of the data we want
to actually plot, we need to give an aesthetic. We can do this by declaring 
that **the aesthetic for the x-axis is the day of the week of the birth**.

The column `DOB_WK` gives the day of the week that each birth happened as 
a numeric value 1 (Sunday) through 7 (Saturday). We can tell R that this is 
actually a factor variable by putting the variable name in the function 
`factor`. Putting this all together, we get `geom_bar(aes(x = factor(DOB_WK))`. 
Finally, to add this layer to the initialized graph, we add the `geom` to `ppp` 
with the `+` operator. 

```{r birth_wk_hist}
ppp <- ggplot(births_small) 
ppp + geom_bar(aes(x = factor(DOB_WK)))
```

Doctors are able to delay birth with anti-contraction medications or labor
suppressants called *tocolytics*. That might be one reason we see fewer births 
on day 1 of the week (Sunday) and day 7 (Saturday).

We can get further information from this plot if we add more aesthetics. 
For example, maybe we can fill each bar with different colors corresponding 
to what type of birth it was ("C-section", "Vaginal", or "Unknown"). We can do 
this by just including another `aes` in the geometric object. Start with the 
same initialization, but add a geometric object with the x-axis and also fill 
color defined.

```{r birth_wk_hist_color}
ppp <- ggplot(births_small) 
ppp + geom_bar(aes(x = factor(DOB_WK), fill = DMETH_REC))
```

When we made that plot, we used the default value for the `position` 
argument of the `geom_bar` function. We could have equivalently written the 
code as follows.

```{r birth_wk_hist_stack}
ppp <- ggplot(births_small) 
ppp + geom_bar(aes(x = factor(DOB_WK), fill = DMETH_REC), position = "stack")
```

Another option is to use `position = "dodge"`. Note that this is an 
argument to `geom_bar` and not to `aes`.

```{r birth_wk_hist_dodge}
ppp <- ggplot(births_small) 
ppp + geom_bar(aes(x = factor(DOB_WK), fill = DMETH_REC), position = "dodge")
```

Now we see that about 1/2 as many C-sections were on weekends as there were on 
a typical weekday, whereas there were about 3/4 as many vaginal births on 
weekends as there were on weekdays.

## Facets

Let's continue looking at the birth data on a day to day basis.  We might 
conjecture that older women are less likely to take a *tocolytic* since they 
are more likely to have had other complications already. One way we can 
do this is to "facet" the graph and display day of the week versus women's age.

First, let's make a histogram of the women's ages to get an idea of what the 
distribution looks like.

```{r birth_age_hist}
ppp <- ggplot(births_small) 
ppp + geom_histogram(aes(x = MAGER), binwidth = 1)
```

We used the argument `binwidth` to set the width of the bins in 
this histogram (here `binwidth = 1` corresponds to 1 year).

Using the grammar of graphics, it is easy to facet this single graph to make 
multiple graphs along another dimension of the data. In this case, we're 
interested in breaking this up along the dimension of day of birth ``DOB_WK``. 
We will add this facet with the command  `facet_grid` or `facet_wrap`.
In `facet_grid`, the argument we use is a formula with the rows (of the 
tabular display) on the left hand side and the columns (of the tabular display) 
on the right hand side (RHS). 

A formula is created in R with the tilde operator `~`.  A dot in the formula 
is used to indicate there should be no faceting on this dimension (either row 
or column). The formula can also be provided as a string instead of a classical 
formula object. In `facet_wrap`, we have the same sort of argument, but we 
only include a RHS of the formula. We'll use both of them in an example so you
can see the difference.

Now let's look at this faceting on that variable. Again, we will use the `+`
operator. Here, we also see that we can save the geometric objects in the plot
and just add facets at the end.

```{r birth_age_facet_wrap}
ppph <- ggplot(births_small) + 
   geom_histogram(aes(x = MAGER, fill = DMETH_REC), binwidth = 1)
ppph + facet_wrap( ~ DOB_WK)
```


```{r birth_age_facet_grid, fig.height=7, fig.width=7}
ppph + facet_grid(DOB_WK ~ SEX)
```


What is the difference between ``facet_wrap`` and ``facet_grid``?

Here is an interesting perspective of the data 
(we use ``dplyr::filter`` to exclude the record where the delivery method 
is Unknown).

```{r birth_age_facet_color}
ggplot(dplyr::filter(births, !(DMETH_REC %in% "Unknown"))) +
  geom_histogram(aes(x = MAGER, fill = factor(TBO_REC)), binwidth = 1) +
  facet_grid(WEEKEND ~ DMETH_REC, scale="free_y", drop = TRUE) +
  geom_vline(xintercept = seq(15, 45, by=5), alpha=0.2, color="white") +
  labs(title = "Births in USA 2006", fill="Birth\nOrder")
```

## Statistics

It's often useful to transform your data before plotting, and that's what 
statistical transformations do.

Here we look at the histogram of mother's ages as before, but we also **add 
a density estimate to it**.

```{r birth_age_dens}
ggplot(births_small, aes(x = MAGER)) + 
  geom_histogram(aes(y = ..density..), binwidth = 1, fill = "grey", col = "black") +
  stat_density(col = "red", fill = NA)
```

Maybe we want to compare this to a lognormal distribution's density.

```{r birth_age_dens_normal}
ggplot(births_small, aes(x = MAGER)) + 
  geom_histogram(aes(y = ..density..), binwidth = 1, fill="grey", col="black") +
  stat_density(col="red", fill=NA) +
  stat_function(fun = dlnorm, col = "blue", 
                args = list(mean = mean(log(births_small$MAGER)),
                             sd =    sd(log(births_small$MAGER))))
```

You can add any function using the `stat_function`.

In this example, we will plot the birth weight in grams (DBWT) versus weight
gain (WTGAIN) by mother (which seems to be in pounds). Unfortunately, we need 
to be a little more careful about this when there are NAs in the data.

```{r birth_weight_scatter}
ppp2 <- ggplot(dplyr::filter(births_small, !is.na(WTGAIN) & !is.na(DBWT)), 
               aes(x = WTGAIN, y = DBWT)) +
  labs(x = "Weight Gain by Mother", y = "Birth Weight in Grams")
ppp2 + geom_point()
```

When we plot this data by itself, it is not clear what is going on 
-- there are many points plotted on top of each other and it just looks messy.

One way to transform this data with a ``stat_bin2d()`` in ``ggplot2`` 
is by **binning the points** as below.

```{r birth_weight_bin2d}
ppp2 + stat_bin2d()
```

See that the color axis reports counts. That means we count the total number 
of observations to fall within a box, i.e. we are effectively **generating
a 2D density plot**. This plot seems to indicate that the joint distribution 
of weight gain by mother and birth weight in grams is unimodal.

Instead of just making counts though, we can compute other quantities
(evaluate functions other than count) within each of the 2d bins on other two 
columns of `births_small`. For example, let's look at the median number of 
weeks of gestation for each of the bins.

```{r birth_weight_bin2d_median}
ppp2 + stat_summary_2d(aes(z = ESTGEST), fun = median) + 
  labs(title = "median number of weeks of gestation")
```

If we want to do a regression, we can include this automatically. By default, 
`ggplot2` uses "locally weighted scatterplot smoothing" (loess) regression 
for data sets with < 1000 points and "generalized additive models" 
(GAM) for >= 1000 points.

```{r birth_weight_lm}
# here we force a linear model fit
ppp2 + geom_point() + stat_smooth(method = lm) 
```

By changing the color aesthetics of the modelled line, we can fit separate 
models to the corresponding subsets of the data and compare them. 
For example, let's look at a smoothed fit (GAM) of birthweight versus 
weight gain by mother separated out by baby boys and girls.

```{r birth_weight_gams}
ppp2 + geom_point() + stat_smooth(aes(col = SEX))
```


Finally, let's look at hexagonal bins as a visually more attractive alternative 
to the rectangular bins used so far

```{r birth_weight_hex}
ppp2 + geom_hex(bins = 30) + stat_smooth(aes(col = SEX))
```

## Scales

Scales control the mapping between data and aesthetics. Going back to the 2d 
distribution of birthweight versus weight gain by mothers, it is difficult to 
see what is going on except that there is a dense region and a less dense 
region. If we take the square root of the number of births per region, then 
we can see that there is a smooth transition between the high density area and 
the low density area.

```{r birth_weight_bin2d_scale}
# we repeat again the code above so you don't gave to scroll to look at it.
ppp2 <- ggplot(dplyr::filter(births_small, !is.na(WTGAIN) & !is.na(DBWT)), 
               aes(x = WTGAIN, y = DBWT)) +
  labs(x = "Weight Gain by Mother", y = "Birth Weight in Grams")

ppp2 + stat_bin2d() + scale_fill_gradient(trans = "sqrt")
```

See what happens when you change the trans to ``log10``.

Sometimes, we might want to change the scale of the vertical axis.
According to Wikipedia's page on "Fetal Viability", there is a 50% chance 
of viability at 24 weeks of gestation. We will repeat the plot above
with birth weight on a log scale, so we get better separation of the mean 
weeks of gestation.

```{r birth_weight_bin2d_trans_estgest}
ppp2 + stat_summary_2d(aes(z = ESTGEST), fun = mean) + 
  scale_y_log10(limits = 10^c(2, 4)) +
  scale_fill_gradient2(midpoint = 24) +
  labs(title="mean number of weeks of gestation")
```

Now let's look at only the quadruplets in the full dataset (there are 39 such 
 observations). We want to include lots of variables such as number of
prenatal visits (UPREVIS), the mother's age (MAGER), the estimated weeks of 
gestation (ESTGEST), the delivery method (DMETH_REC), and the mother's education
level (DMEDUC).

```{r birth_weight_quads}
ppp3 <- ggplot(dplyr::filter(births, DPLURAL == "4 Quadruplet"), 
               aes(x = UPREVIS, y = MAGER)) + 
  geom_point(aes(size = ESTGEST, shape = DMETH_REC, col = DMEDUC)) +
  stat_smooth(aes(col = DMETH_REC), method = "lm")
ppp3
ppp3 + scale_size(range=c(3, 6)) + scale_color_brewer(palette = "Set1") +
  scale_shape(solid = FALSE)
```

Note that, while shapes are discrete, colors and sizes can both be scaled 
continuously.

## Quantile-quantile plot

Remember that in the the power calculations section from the last lab,
we defined a statistic estimating the deviation of the
observed to expected frequencies of nucleotides. The code for generating
the statistics assuming it came from the null distribution (i.e.
when all 4 nucleotides are evenly likely).


```{r}
oestat = function(o, e){
  sum( (e-o)^2/e )
}

set.seed(1)
B = 10000
# here we pick an arbitrary length / not the same as for Celegans
n = 2847
expected = rep(n/4 ,4)
oenull = replicate(
  B, oestat(e=expected, o=rmultinom(1,size = n, prob = rep(1/4,4))))

```

Now, we can estimate the null distribution of this statistics, by
plotting a histogram of the generated values:

```{r}
ggplot(data.frame(null_stats = oenull)) +
  geom_histogram(aes(x = null_stats), bins = 100, boundary=0)
```

We compare the distribution of this statistic to
$Chi^2_3$ (df = 1 * (4-1) = 3) using a q-q plot:


```{r}
 ggplot(data.frame(stat = oenull), aes(sample = stat)) +
   stat_qq(distribution = stats::qchisq, dparams = list(df = 3)) +
   stat_qq_line(distribution = stats::qchisq, dparams = list(df = 3)) 
```



## Parathyroid Example: to complete for Thursday

We will use the ``parathyroidGenesSE`` package in ``R``.  Load the data and 
read the experimental information and the abstract.

```{r para_data, message=FALSE, warning=FALSE}
library("parathyroidSE")
library("EnsDb.Hsapiens.v86")

data("parathyroidGenesSE", package = "parathyroidSE")
metadata(parathyroidGenesSE)$MIAME 
abstract(metadata(parathyroidGenesSE)$MIAME)
```

Parathyroid adenoma (http://en.wikipedia.org/wiki/Parathyroid_adenoma) is 
a benign tumor of the parathyroid gland. The abstract tells us that some 
interesting genes to look at are the following:

* Estrogen related genes: ESR1, ESR2.
* Parathyroid related genes: CASR, VDR, JUN, CALR, ORAI2.

Let's put them in a table.

```{r genes}
genes <- read.csv(textConnection(
  "name, group
   ESR1,  estrogen
   ESR2,  estrogen
   CASR,  parathyroid
   VDR,   parathyroid
   JUN,   parathyroid
   CALR,  parathyroid
   ORAI2, parathyroid"), 
  stringsAsFactors = FALSE, strip.white = TRUE)
```

In the ``parathyroidGenesSE`` object, the features are labeled
with Ensembl gene identifiers, so let's use a Bioconductor package
to find the corresponding IDs.

```{r EnsDb}
ens <- ensembldb::select(EnsDb.Hsapiens.v86,
  keys = list(GenenameFilter(genes$name), 
              TxBiotypeFilter("protein_coding")),
  columns = c("GENEID", "GENENAME"))
ens <- 
  dplyr::filter(ens, GENEID %in% rownames(parathyroidGenesSE)) %>%
  mutate(group = genes$group[match(GENENAME, genes$name)])

ens
```

Make the table of gene counts, add the patient info:

```{r para_cut}
countData <- assay( parathyroidGenesSE ) 
gene.counts <- t(countData[ens$GENEID, ])
colnames(gene.counts) <- ens$GENENAME
dat <- cbind(data.frame(colData( parathyroidGenesSE)), data.frame(gene.counts))
head(dat)
```

Plot one of the estrogen related gene's counts (ESR1) with 
plot aesthetics and faceting to separate patients, treatments and times.

```{r para_plot, tidy=FALSE}
ggplot(dat, aes(col = patient, x = treatment, y = ESR1)) +
  geom_point(size = 3) + 
  facet_grid( . ~ time)
```

## Questions: to complete for Thursday

Answer the following questions.

From the plot of the parathyroid data, answer the following. 

**Quiz question 1** : How many patients are there?

**Quiz question 2** : How many time points are there?
  
**Quiz question 3** : There were 3 treatments: "Control", "DPN", and "OHT". 
How many measurements were taken from patient 2 under the DPN treatment?
  
Make your own plot of VDR versus CASR.  (That is CASR, not CALR).

**Quiz question 4** : Which patient has the highest recorded level of CASR?

**Quiz question 5** : Which of the following pairs of patients seem to be well
separated in this plot (i.e., for which two patients you can draw a line on 
the plot that perfectly separates them)?

**Quiz question 6** : You need to know which pairs of patients are well 
separated with respect to the genes VDR and CASR (i.e., you can draw a 
line on the plot that perfectly separates the patients). Which of the 
following methods can help you visualize this?
  
**Quiz question 7** : Which patient looks different from the other three when 
you plot VDR versus ORAI2?

**Quiz question 8** : Plot ORAI2 versus JUN. Which can you separate best?

**Quiz question 9** : Plot CASR versus (ESR1 + ESR2). Fit a separate 
linear model for each treatment (Control, DPN, OHT). Which linear models 
are increasing?

**Quiz question 10** : What is the maximum number of shapes that you are 
allowed to use in ggplot2 by default?
  
**Quiz question 11** : Write the name of the function that you could use to 
make more than the maximum number of default shapes allowed. Hint: this function 
has "values" as one of the arguments ____(..., values = (...)).

**Quiz question 12** : What do Themes do in ggplot2?
  
## CpG Islands: to complete for Thursday

Let's walk through a last example.

```{r setup2, message = FALSE, warning = FALSE, results = 'hide'}
pkgs_needed <- c("BSgenome.Hsapiens.UCSC.hg19", "Gviz")
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  BiocManager::install(letsinstall)
}
```

This is additional material, which wasn't covered in class but is 
included in the textbook. You will benefit from completing this part.

GC content is the percentage of the genome (or DNA fragment) that is "G" or "C".
To compute the GC content, we count the occurrences of the "G" and "C" alphabets,
and divide by the length of the string in question.

We will be using data from chr8 of the human genome version 19 from the UCSC 
genome repository. 

A genomic window is defined as a CpG island if its GC content>50% and 
observed-to-expected CG ratio>0.6. CpG islands are said to mark important 
regions in the genome because over 65% of gene promoter regions can be found 
in with CpG islands.

We want to look at this for the Human Chromosome 8. We can use Bioconductor 
to get the data with the command `BiocManager::install("BSgenome.Hsapiens.UCSC.hg19")`.

```{r}
library(BSgenome.Hsapiens.UCSC.hg19)
chr8  =  Hsapiens$chr8
cpg_url <- url("http://web.stanford.edu/class/bios221/data/model-based-cpg-islands-hg19.txt")
CpGtab <- read.table(cpg_url, header=T)
nrow(CpGtab)
```

Then we retain only the start and stop positions for chromosome 8 and 
convert into an IRanges object:



```{r}
irCpG = with(dplyr::filter(CpGtab, chr == "chr8"),
         IRanges(start = start, end = end))
```

We create the biological context with the next line. 
The “I” in IRanges stands for “interval”; the “G” in GRanges for “genomic”.

```{r}
grCpG = GRanges(ranges = irCpG, seqnames = "chr8", strand = "+")
genome(grCpG) = "hg19"
```

```{r}
library("Gviz") # you might need to install this from Bioconductor
ideo = IdeogramTrack(genome = "hg19", chromosome = "chr8")
plotTracks(
  list(GenomeAxisTrack(),
    AnnotationTrack(grCpG, name = "CpG"), ideo),
    from = 2200000, to = 5800000,
    shape = "box", fill = "#006400", stacking = "dense")
```


Views on the chromosome sequence correspond to the CpG islands, irCpG, and to
the regions in between (gaps(irCpG)). 

```{r}
CGIview    = Views(unmasked(Hsapiens$chr8), irCpG)
NonCGIview = Views(unmasked(Hsapiens$chr8), gaps(irCpG))
```

We compute transition counts in CpG islands and non-islands using the data.


```{r}
seqCGI      = as(CGIview, "DNAStringSet")
seqNonCGI   = as(NonCGIview, "DNAStringSet")
dinucCpG    = sapply(seqCGI, dinucleotideFrequency)
dinucNonCpG = sapply(seqNonCGI, dinucleotideFrequency)
dinucNonCpG[, 1]
```


```{r}
NonICounts = rowSums(dinucNonCpG)
IslCounts  = rowSums(dinucCpG)
```


For a four state Markov chain as we have, we define the transition matrix as 
a matrix where the rows are the from state and the columns are the to state.


```{r}
TI  = matrix( IslCounts, ncol = 4, byrow = TRUE)
TnI = matrix(NonICounts, ncol = 4, byrow = TRUE)
dimnames(TI) = dimnames(TnI) =
  list(c("A", "C", "G", "T"), c("A", "C", "G", "T"))
```

We use the counts of numbers of transitions of each type to compute frequencies 
and put them into two matrices.

```{r}
MI = TI /rowSums(TI)
MI
```


```{r}
MN = TnI / rowSums(TnI)
MN
```

Are the relative frequencies of the different nucleotides different in
CpG islands compared to elsewhere ?

```{r}
freqIsl=alphabetFrequency(seqCGI,baseOnly=TRUE,collapse=TRUE)[1:4]
freqIsl / sum(freqIsl)
```


```{r}
freqNon=alphabetFrequency(seqNonCGI,baseOnly=TRUE,collapse=TRUE)[1:4]
freqNon / sum(freqNon)
```

This shows an inverse pattern: in the CpG islands, C and G have frequencies
around 0.32, whereas in the non-CpG islands, we have A and T that have 
frequencies around 0.30.

But how can we decide if a new sequence comes from a CpG island?
What is the probability it belongs to a CpG island compared to somewhere else?
We compute a score based on what is called the odds ratio, i.e
$\log( $P(sequence | island)/P(sequence| nonisland) $)$. See the textbook
for details as to how this is evaluated.

```{r}
alpha = log((freqIsl/sum(freqIsl)) / (freqNon/sum(freqNon)))
beta  = log(MI / MN)

scorefun = function(x) {
  s = unlist(strsplit(x, ""))
  score = alpha[s[1]]
  if (length(s) >= 2)
    for (j in 2:length(s))
      score = score + beta[s[j-1], s[j]]
  score
}

x = "ACGTTATACTACG"
scorefun(x)
```


In the code below, we pick sequences of length len = 100 out of the 2855 sequences in the seqCGI object, and then out of the 2854 sequences in the seqNonCGI object (each of them is a DNAStringSet). In the first three lines of the generateRandomScores function, we drop sequences that contain any letters other than A, C, T, G; such as “.” (a character used for undefined nucleotides). Among the remaining sequences, we sample with probabilities proportional to their length minus len and then pick subsequences of length len out of them. The start points of the subsequences are sampled uniformly, with the constraint that the subsequences have to fit in.

```{r}
generateRandomScores = function(s, len = 100, B = 1000) {
  alphFreq = alphabetFrequency(s)
  isGoodSeq = rowSums(alphFreq[, 5:ncol(alphFreq)]) == 0
  s = s[isGoodSeq]
  slen = sapply(s, length)
  prob = pmax(slen - len, 0)
  prob = prob / sum(prob)
  idx  = sample(length(s), B, replace = TRUE, prob = prob)
  ssmp = s[idx]
  start = sapply(ssmp, function(x) sample(length(x) - len, 1))
  scores = sapply(seq_len(B), function(i)
    scorefun(as.character(ssmp[[i]][start[i]+(1:len)]))
  )
  scores / len
}
scoresCGI    = generateRandomScores(seqCGI)
scoresNonCGI = generateRandomScores(seqNonCGI)
```


Now, we can use `ggplot()` to view the distribution of computed scores
for the two distributions:


```{r}
df <- rbind(
  data.frame(region = "cgi", score = scoresCGI),
  data.frame(region = "ncgi", score = scoresNonCGI)
)

ggplot(df, aes(x = score, fill = region)) +
  geom_histogram(alpha = 0.5, color = "black", position = "identity")
```



## In-class exercise

We will walk through the following exercise in class. Feel free to have a look beforehand, but we will solve it in groups during the Thursday session.





