---
title: "Graphics Solutions"
author: "bioX"
date: "07/16/2020"

output: 
  html_document:
    toc: true
    toc_float: true
---


## Load data

```{r para_data, message=FALSE, warning=FALSE}
library("parathyroidSE")
library("EnsDb.Hsapiens.v86")
library("dplyr")
library("ggplot2")

data("parathyroidGenesSE", package = "parathyroidSE")

genes <- read.csv(textConnection(
  "name, group
   ESR1,  estrogen
   ESR2,  estrogen
   CASR,  parathyroid
   VDR,   parathyroid
   JUN,   parathyroid
   CALR,  parathyroid
   ORAI2, parathyroid"), 
  stringsAsFactors = FALSE, strip.white = TRUE)

ens <- ensembldb::select(EnsDb.Hsapiens.v86,
  keys = list(GenenameFilter(genes$name), 
              TxBiotypeFilter("protein_coding")),
  columns = c("GENEID", "GENENAME"))
ens <- 
  dplyr::filter(ens, GENEID %in% rownames(parathyroidGenesSE)) %>%
  mutate(group = genes$group[match(GENENAME, genes$name)])


countData <- assay( parathyroidGenesSE ) 
gene.counts <- t(countData[ens$GENEID, ])
colnames(gene.counts) <- ens$GENENAME
dat <- cbind(data.frame(colData( parathyroidGenesSE)), data.frame(gene.counts))
head(dat)
```

## Initial plot

Plot one of the estrogen related gene's counts (ESR1) with 
plot aesthetics and faceting to separate patients, treatments and times.

```{r para_plot, tidy=FALSE}
ggplot(dat, aes(col = patient, x = treatment, y = ESR1)) +
  geom_point(size = 3) + 
  facet_grid( . ~ time)
```


## Quiz Answers

From the plot of the parathyroid data, answer the following. 

**Quiz question 1** : How many patients are there?

**Answer:** There are four patients.

**Quiz question 2** : How many time points are there?
  
**Answer:** There are two timepoints (24h, 48h).


**Quiz question 3** : There were 3 treatments: "Control", "DPN", and "OHT". 
How many measurements were taken from patient 2 under the DPN treatment?
  
**Answer:** There were three measurements taken for patient 2 (green) under 
DPN treament.
  
**Question**: Make your own plot of VDR versus CASR.  (That is CASR, not CALR).

**Answer**:

```{r, fig.width=8, fig.height=4}
ggplot(dat, aes( x = VDR, y = CASR)) +
    geom_point(aes(fill = patient, shape = treatment, alpha=time), size = 3) +
    scale_alpha_manual(values = c(1.0, 0.7)) +
    scale_shape_manual(values = c(21, 22, 24)) +
    guides(fill = guide_legend(override.aes=list(shape=21))) +
    facet_grid( . ~ time)
```

**Quiz question 4** : Which patient has the highest recorded level of CASR?

**Answer:** Patient 2.

**Quiz question 5** : Which of the following pairs of patients seem to be well
separated in this plot (i.e., for which two patients you can draw a line on 
the plot that perfectly separates them)?

**Answer:** 
Patient 1 and patient 2, also
Patient 1 and patient 3.

**Quiz question 6** : You need to know which pairs of patients are well 
separated with respect to the genes VDR and CASR (i.e., you can draw a 
line on the plot that perfectly separates the patients). Which of the 
following methods can help you visualize this?
  
**Answer:** 
Plot VDR versus CASR, and change the shape of the point according to the patient. 
Plot VDR versus CASR, and color the points according to the patient. 
  
  
**Quiz question 7** : Which patient looks different from the other three when 
you plot VDR versus ORAI2?

**Answer:** Patient 2.

```{r, fig.width=8, fig.height=4}
ggplot(dat, aes( x = VDR, y = ORAI2)) +
    geom_point(aes(fill = patient, shape = treatment, alpha=time), size = 3) +
    scale_alpha_manual(values = c(1.0, 0.7)) +
    scale_shape_manual(values = c(21, 22, 24)) +
    guides(fill = guide_legend(override.aes=list(shape=21))) +
    facet_grid( . ~ time)
```

**Quiz question 8** : Plot ORAI2 versus JUN. Which can you separate best?

**Answer:** 24 hours vs. 48 hours.

```{r, fig.width=6, fig.height=5}
ggplot(dat, aes( x = JUN, y = ORAI2)) +
    geom_point(aes(fill = time, shape = patient, alpha=treatment), size = 3) +
    scale_alpha_manual(values = c(1.0, 0.7, 0.5)) +
    scale_shape_manual(values = c(21, 22, 23,24)) +
    guides(fill = guide_legend(override.aes=list(shape=21)))
```

**Quiz question 9** : Plot CASR versus (ESR1 + ESR2). Fit a separate 
linear model for each treatment (Control, DPN, OHT). Which linear models 
are increasing?

**Answer:** DPN and OHT.

```{r, fig.width=8, fig.height=4}
ggplot(dat, aes( y = CASR, x = ESR1 + ESR2)) +
    geom_point(aes(fill = treatment, color=treatment, shape = patient, alpha=time), size = 3) +
    geom_smooth(aes(color = treatment), method = "lm", se=FALSE)+
    scale_alpha_manual(values = c(1.0, 0.6)) +
    scale_shape_manual(values = c(21, 22, 23,24)) +
    guides(fill = guide_legend(override.aes=list(shape=21)))
```


**Quiz question 10** : What is the maximum number of shapes that you are 
allowed to use in ggplot2 by default?
  
**Answer:** 6.
```{r}
df_shape <- data.frame(x=runif(10),y=runif(10), shapes=as.factor(1:10))
ggplot(df_shape, aes(x=x,y=y, shape=shapes)) + geom_point()
```

**Quiz question 11** : Write the name of the function that you could use to 
make more than the maximum number of default shapes allowed. Hint: this function 
has "values" as one of the arguments ____(..., values = (...)).

**Answer:** `scale_shape_manual`.


**Quiz question 12** : What do Themes do in ggplot2?

**Answer:** They control non-data components of the plot. 


## In-class exercise: Customized scatter plot 

You will try to recreate a [plot](https://www.economist.com/sites/default/files/imagecache/1280-width/images/2016/07/blogs/graphic-detail/20160723_woc155_1.png) from an Economist article
showing the relationship between well-being and financial inclusion.

You can find the accompanying article at this [link](http://www.economist.com/blogs/graphicdetail/2016/07/daily-chart-13)


The data for the exercises `EconomistData.csv` can be downloaded from 
the class github repository.

```{r}
library(tidyverse)
url <- paste0("https://raw.githubusercontent.com/cme195/cme195.github.io/",
              "master/assets/data/EconomistData.csv")
dat <- read_csv(url)
head(dat)
```


### Part 1

1. Create a scatter plot similar to the one in the article, where 
the x axis corresponds to percent of people over the age of 15 with a 
bank account (the `Percent.of.15plus.with.bank.account` column)
and the y axis corresponds to the current SEDA score `SEDA.Current.level`.
2. Color all points blue.
3. Color points according to the `Region` variable.
4. Overlay a fitted smoothing trend on top of the scatter plot. Try to
change the span argument in `geom_smooth` to a low value and see what happens.
5. Overlay a regression line on top of the scatter plot 
Hint: use `geom_smooth` with an appropriate method argument. 
6. Facet the previous plot by `Region`.

```{r}
#1. Create a scatter plot with percent of people over the age of 15 with a bank 
p <- ggplot(
  dat, aes(x = Percent.of.15plus.with.bank.account, y = SEDA.Current.level)) 
p + geom_point()
```

```{r}
#2. Color the points in the previous plot blue.
p + geom_point(color = "blue")
```

```{r}
#3. Color the points in the previous plot according to the `Region`.
(p3 <- p + geom_point(aes(color = Region)))
```


```{r}
# 4. Overlay a smoothing line on top of the scatter plot using the default method.
p3 + geom_smooth()
```

```{r}
#4. Changing the span parameter
p3 + geom_smooth(span = 0.2)
```

```{r}
#5. Overlay a smoothing line on top of the scatter plot using the lm method
(p5 <- p3 + geom_smooth(method = "lm"))
```

```{r}
# 6. Facetting plots
p5 + facet_wrap(~ Region)
```


### Part 2: Distribution of categorical variables

1. Generate a bar plot showing the number of countries included in the
dataset from each `Region`.

```{r}
ggplot(dat, aes(x = Region)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r}
dat <- dat %>%
  mutate(reg = reorder(Region, Region, function(x) -length(x)))
barplot <- ggplot(dat, aes(x = reg)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
barplot
```

2. Rotate the plot so the bars are horizontal

```{r}
barplot + coord_flip()
```



### Part 3: Distribution of continuous variables

1. Create boxplots of SEDA scores, `SEDA.Current.level` separately 
for each `Region`.
2. Overlay points on top of the box plots
3. The points you added are on top of each other, in order to distinguish 
them jitter each point by a little bit in the horizontal direction.
4. Now substitute your boxplot with a violin plot.

```{r}
plt <- ggplot(dat, aes(x = Region, y = SEDA.Current.level)) + 
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
plt + geom_boxplot()  
```

```{r}
plt + geom_boxplot() + geom_point()
```

```{r}
plt + geom_boxplot() + geom_jitter(width = 0.1)
```

```{r}
plt + geom_violin() + geom_jitter(width = 0.1)
```






### Emulating the Economist 'style'

Below, I will show you how to obtain an 'Economist-look' for your scatter plot
in few lines of code. To generate a replicate plot we need to:


1. Change ordering of the regions, by converting `Region` column to a factor.
2. Use seetings for the markers to best match the points on the original 
Economist plot. Note that the points are bigger and have white borders,
and specific fill colors. The following colors match the ones on the plot:
`colors <-  c("#28AADC","#F2583F", "#76C0C1","#24576D", "#248E84","#DCC3AA",
"#96503F")`
3. Change the axes ratio.
4. Change the plot background and theme. Note that `ggthemes` package
has a convenient functions for generating "Economist" style plots, 
e.g. `theme_economist_white()`.
5. Format the legend.
6. Add "Country" labels to the points.
7. Add a title and format the axes.


First, change order of and lables for `Regions`

```{r}
regions <- c("Europe", "Asia", "Oceania", "North America", 
             "Latin America & the Caribbean", "Middle East & North Africa",
             "Sub-Saharan Africa")

# Here we are just modifying labels so that some names are on two lines
region_labels <-  c("Europe", "Asia", "Oceania", "North America",
                    "Latin America & \n the Caribbean", 
                    "Middle East & \n North Africa", "Sub-Saharan \n Africa")
dat <- dat %>%
  mutate(
    Region = as.character(Region),
    Region = factor(Region, levels = regions, labels = region_labels)
  )
```


```{r}
custom_colors <- c("#28AADC","#F2583F", "#76C0C1","#24576D", "#248E84",
                   "#DCC3AA","#96503F")
p <- ggplot(
  dat, aes(Percent.of.15plus.with.bank.account, SEDA.Current.level)) +
  geom_point(aes(fill = Region), color = "white", size = 4, pch = 21) +
  geom_smooth(method = "lm", se = FALSE, col = "black", size = 0.5) +
  scale_fill_manual(name = "", values = custom_colors) +
  coord_fixed(ratio = 0.4) +
  scale_x_continuous(name = "% of people aged 15+ with bank account, 2014",
                     limits = c(0, 100),
                     breaks = seq(0, 100, by = 20)) +
  scale_y_continuous(name = "SEDA Score, 100=maximum",
                     limits = c(0, 100),
                     breaks = seq(0, 100, by = 20)) +
  labs(title="Laughing all the way to the bank",
       subtitle="Well-being and financial inclusion* \n 2014-15")
```

```{r}
p
```



To change the background and theme to match the 'Economist style',
you can install the 
[`ggthemes`](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html) 
package that implements the themes from:

* Base graphics
* Tableau
* Excel
* Stata
* Economist
* Wall Street Journal
* Edward Tufte
* Nate Silver's Fivethirtyeight
* etc.

```{r}
#install.packages("ggthemes")
library(ggthemes)
(p <- p + theme_economist_white(gray_bg = FALSE))
```

Format the legend

```{r, fig.width=7, fig.height=5}
p + theme(
  text = element_text(color = "grey35", size = 11),
  legend.text = element_text(size = 10),
  legend.position = c(0.72, 1.12),   
  legend.direction = "horizontal") +
  guides(fill = guide_legend(ncol = 4, byrow = FALSE))
```

### Add point labels

```{r}
# Choose a subset of countries
pointsToLabel <- c(
  "Yemen", "Iraq", "Egypt", "Jordan", "Chad", "Congo", "Angola", "Albania",
  "Zimbabwe", "Uganda", "Nigeria", "Uruguay", "Kazakhstan", "India", "Turkey",
  "South Africa", "Kenya", "Russia", "Brazil", "Chile", "Saudi Arabia", 
  "Poland", "China", "Serbia", "United States", "United Kingdom")
```

```{r, fig.width=7, fig.height=5}
# install.packages("ggrepel")
library(ggrepel)
(p <- p + 
    geom_text_repel(
      aes(label = Country), color = "grey20",
      data = dat %>% filter(Country %in% pointsToLabel),
      force = 15))
```


## In-Class 2


```{r}
library(readr)
url <- "http://cdiac.ess-dive.lbl.gov/ftp/ndp030/CSV-FILES/nation.1751_2014.csv"
emissions <- read_csv(url)
emissions <- emissions[-(1:3), ]
emissions[emissions == "."] <- NA
emissions <- type_convert(emissions)
emissions
```

a. Use  `dplyr` functions to compute the total yearly $CO_2$ emissions (column `Total.CO2.emissions.from.fossil.fuels.and.cement.production..thousand.metric.
tons.of.C.`) summed over all countries (the world total $CO_2$ emission). 
Use the dataset to plot the World's yearly $CO_2$ emission in Gt.

```{r}
yearlyCO2 <- emissions %>% 
    group_by(Year) %>%
    summarise(total_CO2 = sum(`Total CO2 emissions from fossil-fuels and cement production (thousand metric tons of C)`))
yearlyCO2
```


b. Find the top 10 countries with highest emission after year 2000 (including 
2000).

```{r}
top10 <- emissions %>% 
    filter(Year >= 2000) %>%
    group_by(Nation) %>%
    summarise(CountryCO2Post200 = sum(`Total CO2 emissions from fossil-fuels and cement production (thousand metric tons of C)`)) %>%
    top_n(10) %>%
    .[["Nation"]]
```

Plot the yearly total CO2 emissions of these top 10 countries
with a different color for each country. Use billion tonnes (Gt) units, 
i.e. divide the total emissions by 10^6.


```{r, fig.width=8, fig.height=4}
ggplot(
    emissions %>% 
        filter(Nation %in% top10),
    aes(x = Year, y = `Total CO2 emissions from fossil-fuels and cement production (thousand metric tons of C)`/1e6)) +
    geom_line(aes(color = Nation)) +
    ylab("Total CO2 emission in Gt") +
    theme_bw()
```


c. Use `geom_area()` to generate a plot similar to the one you generated
above but, with emission levels stacked on top
of each other (summing to the total for the ten countries)
with areas colored by countries.

```{r}
df <- emissions %>% filter(Nation %in% top10)
ordered_countries <- df %>% 
    filter(Year == max(Year)) %>% 
    arrange(`Total CO2 emissions from fossil-fuels and cement production (thousand metric tons of C)`) %>% .[["Nation"]]
df <- df %>%
    mutate(Nation = factor(Nation, levels = ordered_countries))
```



```{r, fig.width=8, fig.height=4}
library(RColorBrewer)
colors10 <- colorRampPalette(brewer.pal(9, "Set2"))(10)
ggplot(
    df,
    aes(x = Year, y = `Total CO2 emissions from fossil-fuels and cement production (thousand metric tons of C)`/1e6)) +
    geom_area(aes(fill = Nation), position = "stack") +
    ylab("Total CO2 emission in Gt") +
    scale_fill_manual(values = colors10) +
    theme_bw()
```



