---
title: "Exploratory Data Analysis Exercise"
author: "Yuwei Cheng"
date: "2023-06-13"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

## Motivation

In this lab, we are going to conduct exploratory data analysis for resale house price in Beijing, China.
There are several reasons for me to choose this dataset for you to practice. First, this is a raw dataset fetched from an online website. It contains many missing values, corrupted entries, and wrong data type encoding, which serves as a good practice for your data cleaning skills. Second, this dataset has various data type, for example, categorical variables, continuous variables, strings, links, and etc. This requires different data visualization skills. Finally, this is a large dataset that contains rich information and interesting patterns (more than 20 variables and around 300K records).       

## Feature Description and Associated Data Type
- url: the url which fetches the data(character)
- id: the id of transaction(character)
- Lng: and Lat coordinates, using the BD09 protocol. (numerical)
- Cid: community id (numerical)
- tradeTime: the time of transaction (dates)
- DOM: active days on market. (integer)
- followers: the number of people follow the transaction. (integer)
- totalPrice: the total price (numerical, in 10K, unit RMB)
- price: the average price by square (numerical, unit RMB)
- square: the square of house (numerical)
- livingRoom: the number of living room (character)
- drawingRoom: the number of drawing room (character)
- kitchen: the number of kitchen (integer)
- bathroom the number of bathroom (integer)
- floor: the height of the house (integer)
- buildingType: including tower (1) , bungalow (2)，combination of plate and tower (3), plate (4) (categorical)
- constructionTime: the time of construction (numerical)
- renovationCondition: including other (1), rough(2),Simplicity (3), hardcover (4) (categorical)
- buildingStructure: including unknown (1), mixed(2), brick and wood (3), brick and concrete (4), steel (5) and steel-concrete composite (6). (categorical)
- ladderRatio: the proportion between number of residents on the same floor and number of elevator of ladder. It describes how many ladders a resident have on average. (numerical)
- elevator have (1) or not have elevator (0) (binary)
- fiveYearsProperty: if the owner have the property for less than 5 years (binary)
- subway: if there is subway nearby (1) else (0) (binary)

# Explorartory Data Analysis

## Data Cleaning

### Install and Load Packages
```{r, warning=FALSE, message=FALSE}
# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("ggpubr")
# install.packages("ggcorrplot")
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(ggpubr)
library(knitr)
library(ggcorrplot)
```


### Load Data

Define your home path and load data from your home path

```{r}
# shortcut ctl + shift + enter to run the cell
# shortcut ctl + alt + i to insert a new cell

###
# home_path = 
###
data = read.csv(paste0(home_path, "house_price.csv"),  # data path
                sep = ",",
                stringsAsFactors = FALSE, 
                fileEncoding = "latin1")

# Extract the number of floor as numerical variable
data$floor = as.numeric(sapply(data$floor, function(x) strsplit(x,' ')[[1]][2]))
```


### Generate Data Summary and Answer Questions


```{r}
###
# Enter your code here

###
```


#### Q1
How many rows in this dataset? How many columns in this dataset?

#### Q2
Which variables contain missing value?


### Remove url, id, Cid, DOM from this dataset then Remove Missing Values

```{r}
###
# Enter your code here

###
```


### Convert tradeTime from String to Date

```{r}
###
# Enter your code here

###
```

### Create a new variable year  

```{r}
###
# Enter your code here

###
```


### Convert livingRoom, drawingRoom, bathRoom, constructionTime from string to numeric
```{r}
# data$livingRoom = as.numeric(data$livingRoom)
###
# Enter your code here

###
data = data %>% drop_na()  # Remove NAs introduced by coercion
```


### Add Catergory Labels for Building Type

```{r}
makeBuildingType = function(x)
{
  if(x==1){return('Tower')}
  else if(x==2){return('Bungalow')}
  else if(x==3){return('Mix_plate_tower')}
  else if(x==4){return('plate')}
}

data$buildingType = sapply(data$buildingStructure, makeBuildingType)
data$buildingType = as.factor(data$buildingType)
```

### Add Cateory Labels for Renovation Condition and Building Structure

```{r}
###
# Enter your code here

###
```


### Convert elevator, fiveYearsProperty, subway, district from numeric to categorical variable

```{r}
###
# Enter your code here

###
```


### Remove record which contains wrong values

#### Q3 Which record do you think contains wrong value? Remove that record.

Hint: what is the meaning of ladder ratio?
```{r}
###
# Enter your code here

###
```


#### Q4 How many records and features do you have after data cleaning?

## Data Visualization

### Q5 Plot the Average Resale House Price Per Square After 2011 by District with Following Instructions

1. Label the figure title as "Average Resale House Price Per Square By District Over Time"
2. The figure should contain 13 lines with different colors. Each line represents one district.
3. Label y axis as "Average Price"
4. Set y axis limit as (10000, 110000)
5. Question: what is the general trend of the average price per square over time? which district has the highese average price per square?

```{r}
# Compute average price per square by district over time
data %>% subset(year >= 2012) %>%
  group_by(year, district) %>% summarise(ave_price = mean(price))

###
# Enter your code here
###
```


### Q6 Plot the Geographic Information by District with Following Instructions

1. Set Lng as x axis, Lat as y axis
2. Label the figure title as "Geographic Distribution by District"
3. Question: do you think geographic location is related to the average resale house price? 

```{r}
###
# Enter your code here
###
```


### Q7 Plot the Average Followers After 2011 By District with Following Instructions

1. Label the figure title as "Average Followers By District Over Time"
2. The figure should contain 13 lines with different colors. Each line represents one district.
3. Label x axis as "Year" and label y axis as "Average Followers"
4. Question: Which district has increasing attention for recent years (increasing followers) and why? Hint: you may answer this question based on the information from the previous two plots 

```{r}
###
# Enter your code here
###
```


### Q8 Use Boxplot to Demonstrate Price Quantiles of buildingType, buildingStructure, and renovationCondition with Following Instructions

1. Put three figures into one row.
2. Question: which buildingType tends to have higher average price? Answer the same question for buildingStructure and renovationCondition.

```{r, fig.width=18, fig.height=8}
###
# Enter your code here
###
```


### Q9 Create Density Plot of House Square by Building Type with Following Instructions

1. Use facet_wrap. The figure should contain 6 subplots
2. Question: Which building type do you think tend to be used to build house with smaller size? 

```{r, fig.width=12, fig.height=12}
###
# Enter your code here
###
```


### Q10 On average, how much more (less) you need to pay price per square for house with subway nearby than house without subway? Answer the same question for elevator, and fiveYearsProperty

```{r}
###
# Enter your code here
###
```


### Q11 Display the Top 5 House Construction Time with Associated Counts

```{r}
###
# Enter your code here
###
```


### Q12 Create Correlation Plot for All Numerical Variables in this Dataset 

1. What are the top three most correlated (negative / positive) variables

```{r, fig.width=12, fig.height=12}
###
# Enter your code here
###
```


## Summary

### Q13 From the above analysis, which variables do you think can be used to explain the resale house price?

